---
title: "From Tricky to Trouble-Free Trails: MachLeish Trail Rating System "
author: "Anna Ballou and Jessica Jacinto"
date: "4/15/2019"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 
Link to github repository [^1] 
 
```{r, message = FALSE}
library(tidyverse)
library(sf)
library(macleish)
library(RSQLite)
library(leaflet)
library(readr)
```

**Problem Statement:**
We are attempting to rate the trails at MacLeish by difficulty. Currently, no such rating system for MacLeish exists. In doing so, hikers will be able to choose a trail that best fits their hiking needs. 

**Examine Trails data (group by name and summarize by length) and Contours data: **
```{r, warning = FALSE}
trails <- macleish_layers %>%
  pluck("trails") 

trails <- trails %>%
  mutate(length = st_length(trails)) %>%
  group_by(name) %>%
  summarize(length = sum(length))

contours <- macleish_layers %>%
  pluck("contours_30ft") 
```

**Find the intersections of Each Trail with the Contour lines: **
```{r, warning = FALSE, message = FALSE}
intersection <- st_intersection(
  pluck(macleish_layers, "trails"), 
  pluck(macleish_layers, "contours_30ft"))

western <- filter(intersection, name == "Western Loop")
western_pts <- western %>% 
  st_cast("MULTIPOINT") %>% 
  st_cast("POINT")

vernal <- filter(intersection, name == "Vernal Pool Loop")
vernal_pts <- vernal %>%
  st_cast("MULTIPOINT") %>% 
  st_cast("POINT")

snowmobile <- filter(intersection, name == "Snowmobile Trail")
snowmobile_pts <- snowmobile %>%
  st_cast("MULTIPOINT") %>% 
  st_cast("POINT")

porcupine <- filter(intersection, name == "Porcupine Trail")
porcupine_pts <- porcupine %>%
  st_cast("MULTIPOINT") %>% 
  st_cast("POINT")

poplar_hill <- filter(intersection, name == "Poplar Hill Road")
poplar_hill_pts <- poplar_hill %>%
  st_cast("MULTIPOINT") %>% 
  st_cast("POINT")

eastern <- filter(intersection, name == "Eastern Loop")
eastern_pts <- eastern %>%
  st_cast("MULTIPOINT") %>% 
  st_cast("POINT")
```

**Plot the trails and Contours for future Reference:**
```{r, warning = FALSE}
trail_plot <- leaflet() %>%
  addTiles() %>%
  addPolylines(data = trails, 
               weight = 2, 
               color = "black") %>%
  addPolylines(data = contours,
               weight = 1,
               color = "red") 
trail_plot
```

**Find Elevation Changes**

Note: [^2] 

*Shenandoah's Hiking Difficulty is determined by a numerical rating using the following formula: *

*Elevation Gain x 2 x distance (in miles). The product's square root is the numerical rating.* 

*For example, a 10-mile hike that gains 2,200 feet in elevation:*
*2,200 x 2=4,400 *
*4,400 x 10 = 44,000*
*Square root of 44,000=209.8*


```{r, warning = FALSE}
Shenandoah_western <- western_pts %>%
  select("CONTOUR_FT") %>%
  mutate(change_in_elevation = max(CONTOUR_FT) - min(CONTOUR_FT)) %>%
  group_by(change_in_elevation) %>%
  summarize(mean(CONTOUR_FT))
```
The Western Loop's gain in elevation is 90ft per the above calculation

```{r, warning = FALSE}
Shenandoah_vernal <- vernal_pts %>%
  select("CONTOUR_FT") %>%
  mutate(change_in_elevation = max(CONTOUR_FT) - min(CONTOUR_FT)) %>%
  group_by(change_in_elevation) %>%
  summarize(mean(CONTOUR_FT))
```
The Vernal Pool Loop's gain in elevation is 30ft per the above calculation. 

```{r, warning = FALSE}
Shenandoah_snowmobile <- snowmobile_pts %>%
  select("CONTOUR_FT") %>%
  mutate(change_in_elevation = max(CONTOUR_FT) - min(CONTOUR_FT)) %>%
  group_by(change_in_elevation) %>%
  summarize(mean(CONTOUR_FT))
```
The Snowmobile Trail's gain in elevation is 120ft per the above calculation. 

```{r, warning = FALSE}
Shenandoah_porcupine <- porcupine_pts %>%
  select("CONTOUR_FT") %>%
  mutate(change_in_elevation = max(CONTOUR_FT) - min(CONTOUR_FT)) %>%
  group_by(change_in_elevation) %>%
  summarize(mean(CONTOUR_FT))
```
The Porcupine trail does not change elevation with the 30 ft contour lines (per the above calucation)

```{r, warning = FALSE}
Shenandoah_eastern <- eastern_pts %>%
  select("CONTOUR_FT") %>%
  mutate(change_in_elevation = max(CONTOUR_FT) - min(CONTOUR_FT)) %>%
  group_by(change_in_elevation) %>%
  summarize(mean(CONTOUR_FT))
```

The Eastern Loop's gain in elevation is 30ft per the above calculation. 
```{r, warning = FALSE}
Shenandoah_poplar <- poplar_hill_pts %>%
  select("CONTOUR_FT") %>%
  mutate(change_in_elevation = max(CONTOUR_FT) - min(CONTOUR_FT)) %>%
  group_by(change_in_elevation) %>%
  summarize(mean(CONTOUR_FT)) 
```

The Poplar Hill Road's gain in elevation is 60ft per the above calculation. 


**Finding Shenandoah ranking: **

*Elevation Gain x 2 x distance(miles)*

Western Loop: square-root[(90)(2)(0.8392)] = 15.92

Vernal Pool Trail:
square-root[(30)(2)(0.2240)] = 2.45

Snowmobile Trail: 
square-root[(120)(2)(1.6)]
= 35.05

Porcupine Trail:
= 0 

Eastern Loop:
square-root[(30)(2)(1.2)] 
= 13.15

Poplar Hill Trail:
square-root[(60)(2)(0.65))]
= 10.069

**Create a data.frame with the trails and their Shenandoah rankings**

*Note: Shenandoah rankings involve much more techerous trails, thus the ranges of rankings were adjusted for the RELATIVE difficulty of the MacLeish trails*

```{r, warning = FALSE, message = FALSE}
shenandoah_ranking <- read_csv("shen_rankings.csv")
```

```{r, warning = FALSE, message = FALSE}
ranked_trails <- trails %>%
  full_join(shenandoah_ranking, by = c("name")) %>%
  mutate(rank = ifelse(ranking < 15, "Easy", ifelse(ranking < 25 & ranking > 15, "Moderate", ifelse(ranking > 25 , "Difficult", "Not Ranked"))))
  
```                    

**Create a data graphic that shows the trails ranked by difficulty**
```{r, warning = FALSE}
factpal <- colorFactor(palette = c("red", "green", "orange"), domain = ranked_trails$rank)
```

```{r, warning = FALSE}
trail_ranking_plot <- leaflet() %>%
  addTiles() %>%
  addPolylines(data = ranked_trails, 
               weight = 2.5, 
               color = ~factpal(rank)) %>%
  addLegend(colors = c("green", "orange", "red"), position = c("bottomleft"), labels = c("Easy", "Moderate","Difficult"))
trail_ranking_plot
```


**Next Steps for Macleish**

Now that we know this information it would be great for the people who intend to hike these trails to know it too. Sometimes it is difficult for a person to gauge how long something is just based off of mileage. For most hikes that is really all the information that is given on signage and the rest you can see in brochures or get by word of mouth. It would be great to see this on a map of the grounds so people can visualize what they are getting into. There is also the added possibility of adding the color system to signs on the trails themselves. It wouldn't take much to do either of these things. It may mean having to repaint the signs every few years because the weather here will strip down the paint after a while. The idea of a single map just means there needs to be something protecting it from the weather. That means either putting it on a building that already stands and put some plating around it or creating an outdoor display that can hold it. It would actually be nice to have a few of them around so that people could find there way around nicely. Most parks have structures like these and they seem to stand the test of time. Plus this is a much more eco-friendly alternative to giving every guest a brochure with a map of the whole place.



**Footnotes**

[^1]: https://github.com/aballou16/Macleish-mp3.git
[^2]: https://www.nps.gov/shen/planyourvisit/how-to-determine-hiking-difficulty.htm

